---
title: "Analyze Protein Categories"
subtitle: ""
author: "[Benjamin Dubreuil](https://github.com/benjamin-elusers/)"
home: "https://github.com/benjamin-elusers/yeastomics/"
github: "benjamin-elusers"
mail: "benjamin.dubreuil@weizmann.ac.il"
date: "`r format(Sys.time(), 'Last Modified: %d-%b-%Y %R %p')`"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    highlight: tango
#  html_document: 
    code_download: true
    code_folding: hide
    fig_caption: yes
    keep_md: yes
#    toc: yes
    toc_depth: 3
    toc_float:
      toc_collapsed: true
    number_sections: true
    theme: lumen
  editor_options: 
    chunk_output_type: console
  bibliography: bibliography.bib
  csl: biomed-central.csl
---
<!-- Github Logo -->
<a href="https://github.com/benjamin-elusers/yeastomics/" class="github-corner" aria-label="View source on Github">
  <svg width="80" height="80" viewBox="0 0 250 250" style="fill:#69b3a2; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true">
    <path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path>
    <path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path>
    <path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path>
  </svg>
</a>

<style> 
.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}
@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}
@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE,
                      cache.lazy = FALSE,
                      autodep = TRUE,
                      warning = FALSE)
if('utils' %in% search()){ detach('utils') }
utils = attach(NULL,name = 'utils')
if('funcs' %in% search()){ detach('funcs') }
funcs = attach(NULL,name = 'funcs')

suppressPackageStartupMessages({
  library(tictoc)
  library(tidyverse)
  library(stringr)
  library(hablar)
  
  
  source("src/utils.r")
  source("src/function_annotation.r")
  source("src/function_sequence.r")
  source("src/function_phylogenetic.r")
  source("src/function_analysis.r")
  source("src/function_datalocal.r")
  source("src/function_datapub.r")
})

# turn off annoying messages from dplyr::summarise
options(dplyr.summarise.inform = FALSE)
```

# Evolutionary dataset

```{r evodata}
# EVOLUTIONARY RATE
ER = load.dubreuil2021.data(1) %>% 
     as_tibble() %>%
     dplyr::select(c(starts_with('EVO.'),'PPM','ORF','UNIPROT')) %>%
     hablar::convert(dbl(EVO.FULL), dbl(PPM)) %>% 
     ungroup()

# Fitting mean Evolutionary rate (R) to Protein abundance (E)
init.params =  list(Asym = 3, xmid = 1.5, scal=-1)
nls.fit = nls(EVO.FULL ~ SSlogis(PPM,Asym,xmid,scal),start=init.params,data = ER )
fit.ER = broom::augment_columns(x = nls.fit, data = ER) %>% 
         mutate(SSR = sum(.resid^2), SR = sum(.resid) )

```

# Protein categories

## Loading precomputed data

```{r protcat}
# protein properties (qualitative)
PROP=readRDS(get.last.file("data","proteome-properties"))
PROP_lgl = PROP %>% 
            ungroup() %>% 
            dplyr::select( where(is.logical) )  %>% 
            convert(num(everything())) %>%
            as.matrix
# protein features (quantitative)
FEAT=readRDS(get.last.file("data","proteome-features"))
FEAT_num = FEAT %>%
            ungroup() %>% 
            dplyr::select( where(is.numeric) )

TOT_PROP=ncol(PROP_lgl)
TOT_FEAT=ncol(FEAT_num)
# ER FITTED PROPERTIES
ER.PROP = fit.ER %>% left_join(PROP,by=c('UNIPROT'='UNIPROTKB','ORF'='ORF'))
dim(ER.PROP)
ER.FEAT = fit.ER %>% left_join(FEAT,by=c('UNIPROT'='UNIPROTKB','ORF'='ORF'))
dim(ER.FEAT)
```


There are a total of **`r TOT_PROP`** proteome properties and **`r TOT_FEAT`**
We also gathered **`r TOTAL_FEATURES`** protein features.
## Compute properties similarities

```{r properties-similarity}
# spearman correlation between properties
prop.cor="output/properties-correlation.rds"
PROP_cor = preload(prop.cor,
                   coop::pcor(PROP_lgl), 
                   doing="Computing correlation between properties...")
library(proxy)
tmp = 
tmp[ between(tmp,-0.1,0.1) ] = NA
diag(tmp)=NA

res<-NbClust(PROP_cor, diss=NULL, distance = "binary", min.nc=2, max.nc=100, 
            method = "ward.D2", index = "kl") 
res$All.index
res$Best.nc
table(res$Best.partition)
head(res$Best.partition)
max_cor = apply(abs(tmp),2,max_)
quantile_( max_cor )

uncorrelated = tmp[ which(max_cor < 0.4), which(max_cor < 0.4) ]
colnames( uncorrelated )
pheatmap::pheatmap(tmp, border_color = NA, fontsize = 3,
                   show_colnames = T, show_rownames = F,
                   filename = "output/properties-correlation.pdf",
                   width = 10, height=10
                  )

# jaccard similarity between properties
prop.jac="output/properties-jaccard.rds"
PROP_jac = preload(prop.jac, 
                   dist(PROP_lgl, pairwise = T, method="Jaccard"),
                   doing="Computing Jaccard similarity between properties...") 

# pheatmap::pheatmap(PROP_jac,
#                    show_colnames = F, show_rownames = T,
#                    cluster_cols = F, cluster_rows = F,
#                    clustering_distance_rows = PROP_jac, 
#                    clustering_distance_cols = PROP_jac,
#                    fontsize=5, border_color = NA, silent=T, # cellheight=3, cellwidth=3,
#                    filename="plots/properties-jaccard.pdf", height=30, width=30)
# 
# pheatmap::pheatmap(PROP_cor,
#                    show_colnames = F, show_rownames = T,
#                    cluster_cols = T, cluster_rows = T,
#                    clustering_distance_rows = PROP_cor, 
#                    clustering_distance_cols = PROP_cor,
#                    fontsize=5,border_color = NA, silent=T, # cellheight=3, cellwidth=3,
#                    filename="plots/properties-spearman.pdf", height=30, width=30)


# library(proxy)
# jac.go = dist(x = unigo.m[,-1], method = 'Jaccard',pairwise = TRUE)
# cor.go = cor(unigo.m[,-1], use = 'pairwise',method = 'spearman') # similarity between go term
# #cutree( hclust(as.dist(sim.go)), k=10 )
# pheatmap::pheatmap(1-jac.go,cellheight=3,cellwidth=3,fontsize=4, height=20, width=20,border_col=NA,file='plots/jaccard_dist_go.pdf')
# pheatmap::pheatmap(1-cor.go,cellheight=3,cellwidth=3,fontsize=4, height=20, width=20,border_col=NA,file='plots/correlation_dist_go.pdf')

```


## Residual evolutionary rate
```{r ER}

library(ggplot2)
library(ggthemes)
# Show the fit and the residuals of expression-evolutionary rate
prop_name = grep("essential",colnames(PROP),v=T)
prop_sub = fit.ER %>% filter(!!sym(prop_name))
MR = ggplot(fit.ER, aes(y = EVO.FULL, x = PPM)) +
     geom_abline(slope=0,intercept = mean(fit.ER$EVO.FULL),size=1)+
     geom_point(fill='lightgray',shape=19,alpha=0.3) +
     geom_segment(aes(xend = PPM, yend = .fitted), col='gray',linetype='12',size=0.5) +
     # Overlay the protein sharing a particular properties
     geom_point(data=prop_sub,color='purple',fill=NA,shape=21,size=2)+
     geom_line(mapping=aes(y=.fitted),size=1) +
     xlab('Protein abundance (log10 ppm)') + ylab('Mean ER (full seq.)') +theme_clean() 
MR

# Show residuals of a particular categories 
MR2= ggplot(data=fit.ER, aes(y=.resid,x=PPM)) +
  geom_point(col='lightgray',fill='black',size=2,shape=19) +
  geom_segment(aes(xend = PPM, yend = 0),col='gray',linetype='12',size=0.5) +
  geom_abline(slope=0,intercept = 0,size=1)+
  geom_point(data=prop_sub,color='purple',fill=NA,shape=21,size=2)+
  geom_segment(data=prop_sub,aes(xend = PPM, y=0, yend = .resid),col='purple') +
  xlab('Protein abundance (log10 ppm)') + ylab('Residuals (ER)') +theme_clean() 
MR2


# Show most over/under estimated categories
ER.residuals = fit.ER %>% 
  pivot_longer( cols= starts_with('cat_'), 
                names_to = c('categories','source','property'), 
                names_pattern="cat_(.+)\\.(.+)\\.(.+)",
                values_to = "has_prop") %>%
  filter(has_prop) %>%
  relocate(SGD,GNAME,PNAME,UNIPROT,ORF,categories,source,property,has_prop) %>%
  arrange(categories,source,property,ORF) %>%
  mutate( total.resid = sum(abs(.resid)) ) %>%
  group_by(categories,source, property) %>% 
  summarise(  N=n(),
              N.neg  = sum(.resid<0), 
              N.pos  = sum(.resid>0),
              dN     = N.pos-N.neg,
              rN = N.pos/N.neg,
              SR = sum(.resid),
              SR.log = sign(SR) * log(abs(SR)),
              SR.neg = sum(.resid[.resid<0]), 
              SR.pos = sum(.resid[.resid>0]),
              SR.avg = mean_(.resid), 
           ) %>% 
    distinct() %>% 
  arrange(desc(SR.avg)) 
library(ggplot2)
library(ggfittext)
fig.0 = ggplot(ER.residuals, 
               aes(y = reorder(str_trunc(property,w=50), -SR.log),
                   x = SR.log, fill=source)) +
      geom_bar(stat='identity',position='dodge') + 
      theme_clean() +
      theme( axis.text = element_text(size=3), 
             line = element_line(size=0.1),
             rect = element_rect(size=0.1))
ggsave(fig.0, path='output/', 
       file='barplot-protein-properties-ALL.pdf',
       scale=5,limitsize = F)

fig.1 = fig.0 + 
        facet_wrap(~categories,scales = 'free_y') +
        theme( axis.text = element_text(size=4), 
             line = element_line(size=0.2),
             rect = element_rect(size=0.2))

ggsave(fig.1, 
       path='output/', file='barplot-protein-properties-CATEGORIES.pdf',
       scale=4,limitsize = F)


```
